---
title: 'Gene Set Analysis for GO:0090398 Cellular Senescence'
author: "Jose F Oviedo"
date: "`r Sys.Date()`"
output:
  html_document:
    df_print: paged
  pdf_document: default
---
Cellular senescence is a cell aging process stimulated in response to cellular stress, whereby normal cells lose the ability to divide through irreversible cell cycle arrest. 
    
* regulation (up or down) of cellular senescence is a process that modulates the frequency, rate or extent of cellular senescence. 

* oncogene-induced cell senescence is a cellular senescence process associated with the dismantling of a cell as a response to oncogenic stress, such as the activation of the Ras oncogenic family.
    
* stress-induced premature senescence is a cellular senescence process associated with the dismantling of a cell as a response to environmental factors such as hydrogen peroxide or X-rays.

## setup env

```{r setup}

library(tidyverse)
library(rmarkdown)
library(readxl)
library(knitr)
library(ggpubr)
library(gt)
library(gtExtras)
```

## step 1: read data to env
    
```{r import, message=FALSE, warning=FALSE}
## read in raw read count and DE analysis data frames and the GO class data sets for cellular senescence

SY_read_counts <- read_excel("SY_read_counts.xlsx")

SY_DE_As25_vs_Ctrl_Readcounts <- read_excel("SY_DE_As25_vs_Ctrl_Readcounts.xlsx")

GO_term_summary_pos_reg <- read_excel("cellular_senescence_subcats/GO_term_summary_pos-reg.xlsx")

GO_term_summary_neg_reg <- read_excel("cellular_senescence_subcats/GO_term_summary_neg-reg.xlsx")

GO_term_summary_onco_induced <- read_excel("cellular_senescence_subcats/GO_term_summary_onco-induced.xlsx")

GO_term_summary_stress_induced <- read_excel("cellular_senescence_subcats/GO_term_summary_stress-induced.xlsx")
```

## step 2: data manipulation

```{r data-manipulation-GO-merge}
## merge the GO cellular senescence subcategory data frames

GO_terms_merged <- 
  bind_rows(list(GO_term_summary_pos_reg,
                 GO_term_summary_neg_reg,
                 GO_term_summary_onco_induced,
                 GO_term_summary_stress_induced))
```


```{r data-manipulation-DE-mod1-initialized}
## for practice I selected essential variables to reduce df size and changed the
## 'external_gene_name variable' to 'Symbol' to match the GO_terms_merged variable

DE.mod1 <- SY_DE_As25_vs_Ctrl_Readcounts |>
  rename(Symbol = external_gene_name) |>
  select(c(3,4,7:19))
```


```{r data-manipulation-DE-mod2}
## mutate the DE df for mean

DE.mod2 <- DE.mod1 |>
  mutate(avg_CT_count = (rowSums(DE.mod1[12:15])/4)) |>
  mutate(avg_As_count = (rowSums(DE.mod1[8:11])/4))
```


## step 3: data organization

```{r data-organization}
## filter selected genes from DE.modified$Symbol df using GO_term_summary_pos_reg$Symbol variable
## use the proper join() to only keep the relative GO genes

DE.GO_merged <-
  inner_join(DE.mod2, GO_terms_merged, by = "Symbol") |>
  select(c(1,8:17,19,22))

```

## step 4: down stream analysis

```{r DE-sd}
## sd output is calculated by row for both classes of samples As and CT

CT.sd <- apply(DE.GO_merged[c(6:9)], 1, sd)

As.sd <- apply(DE.GO_merged[c(2:5)], 1, sd)

## now cbind() the array as a column to the DE.GO_merged_complete df

DE.GO_merged_complete <- DE.GO_merged |>
  cbind(CT_sd = CT.sd) |>
  cbind(As_sd = As.sd) |>
  select(1, 10:15)
```

```{r quick-table}
## use gt() for a nice quick table

DE.GO_merged_complete |>
  gt()
```


## step 5: data visualization

```{r plot-avg-CT-DEcount}
## Plot AVG read count and STDDEV for As and CT samples for selected set of genes

pCT <- ggplot(DE.GO_merged_complete,
       aes(Symbol, avg_CT_count, fill = Symbol)) +
  geom_col() +
  labs(title = "Control Group(CT)", subtitle = "Average DE Count Cellular Senescence") +
  xlab("GO Class Genes") +
  ylim(0, 1250)
```

```{r plot-avg-As-DEcount}
## Plot AVG read count and STDDEV for As and CT samples for selected set of genes

pAs <- ggplot(DE.GO_merged_complete,
       aes(Symbol, avg_As_count, fill = Symbol)) +
  geom_col() +
  labs(title = "Treatment Group(As)", subtitle = " Average DE Count Cellular Senescence") +
  xlab("GO Class Genes") +
  ylim(0, 1250)
```

```{r plot-CT-sd}
## Plot AVG read count and STDDEV for As and CT samples for selected set of genes

pCTsd <- ggplot(DE.GO_merged_complete,
       aes(Symbol, CT.sd, fill = Symbol)) +
  geom_col(alpha = 0.6) +
  labs(title = "Control Group(CT)", subtitle = "DE Count Standard Deviation") +
  xlab("GO Class Genes") +
  ylim(0, 300)
```

```{r plot-As-sd}
## Plot AVG read count and STDDEV for As and CT samples for selected set of genes

pAssd <- ggplot(DE.GO_merged_complete,
       aes(Symbol, As.sd, fill = Symbol)) +
  geom_col(alpha = 0.6) +
  labs(title = "Treatment Group(As)", subtitle = "DE Count Standard Deviation") +
  xlab("GO Class Genes") +
  ylim(0, 300)
```

```{r fig1-combined-plots}
## combined single variable plots in one figure 

fig1 <- ggarrange(pCT, pCTsd, pAs, pAssd)

fig1
```

```{r plot-avg-count-vs-sd-CT}

pCTvsSD <- ggplot(DE.GO_merged_complete,
       aes(CT.sd, avg_CT_count)) +
  geom_point(aes(color = Symbol), size = 5) +
  scale_color_discrete(name = "GO Genes") +
  labs(title = "Control Group(CT)", subtitle = "Average DE Count vs STDDEV") +
  ylim(0, 1250) +
  xlim(0, 215)
```


```{r plot-avg-count-vs-sd-As}

pAsvsSD <- ggplot(DE.GO_merged_complete,
       aes(As.sd, avg_As_count)) +
  geom_point(aes(color = Symbol), size = 5) +
  scale_color_discrete(name = "GO Genes") +
  labs(title = "Treatment Group(As)", subtitle = "Average DE Count vs STDDEV by GO Class Gene") +
  ylim(0, 1250) +
  xlim(0, 215)
```


```{r fig2-combined-plots}
## combined multi-variable plots in one figure

fig2 <- ggarrange(pCTvsSD, pAsvsSD, ncol = 1, nrow = 2)

fig2
```