---
title: "How to install addins and set up keyboard shortcuts for them"
author: "Write up by Jose F Oviedo using rstudioaddins tutorial, linked below"
date: "03/09/2023"
output:
  prettydoc::html_pretty:
    theme: cayman
    highlight: github
---

# This is a addins practice tutorial using 'addinexamples' package from CRAN library.

#### Link To Tutorial: [rstudioaddins ](https://rstudio.github.io/rstudioaddins/)

``` {r}
## R SETUP

library(tidyverse)
library(addinexamples)
library(shiny)
library(prettydoc)
library(knitr)
```

Use addins function to quickly show subset of a data frame in a pretty way using 'shiny'

```{r results='hide'}
## use any data set from tidyverse package
## data() to view them

data()
```


```{r}

## I am using the USPersonalExpenditure data frame

USPersonalExpenditure ## then execute the sub set addin from Tools tab


## it is very small data set maybe choose another 

mtcars ## execute addin \
```

The mtcars data frame works much better. Try out some other data frames that are already in your package library and try to remember you have this quick function in your toolbox. Maybe set up a keyboard shortcut for this addin function if you use it often.

The next thing you probably want to do is set up keyboard shortcuts for the addin functions that are used the most, and this is done after selecting Tools>Addins>BrowseAddins and is located at the bottom of the window.

